---
title: "America's Honey Production"
author: "Kasim Helil"
output: html_document
---


# Introduction

Decline in the number of honeybee was a major concern which is a major component to American honey agriculture, according to the National Agricultural Statistics Service(NASS) analysis, In 2006. 

>Large numbers of hives were lost to Colony Collapse Disorder; a phenomenon of disappearing worker bees causing the remaining hive colony to collapse.

There are many speculations to the cause of this disorder points to hive diseases and pesticides harming the pollinators, though no overall consensus has been reached. After twelve years, some industries are observing recovery but the American honey industry is still largely struggling. The U.S. used to locally produce over half the honey it consumes per year. Now, honey mostly comes from overseas, with 350 of the 400 million pounds of honey consumed every year originating from imports. This toutorial provides insight into honey production supply and demand in America by state by analyzing the data from NASS, reporting body for the US Department of Agriculture (USDA) between the years 1998 to 2012.


- How has honey production yield changed from 1998 to 2012?

- Over time, which states produce the most honey? Which produce the least? Which have experienced the most change in honey yield?

- Are there any patterns that can be observed between total honey production and value of production every year? How has value of production, which in some sense could be tied to demand, changed every year?

# Tyding Data

- The following section demonstrates the steps taken to clean and prepare the honeyproduction dataset.
Raw .csv files can be found on this page <https://www.nass.usda.gov/About_NASS/index.php>

The following code chunck will produce 3 datasets with the same type of information on honeybee production and prices. The datasets are for each year interval (1998-2002, 2003-2007, 2008-2012). We will use the read_lines() fuction, becuase the data is messy.


We start by importing the libraries we need to perform the taskes, I personally prefer to keep all the imported libraries together, because it will make the aesthetics look inviting to read than having them in different code chunks. 

```{r library, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(data.table)
library(formattable)
```

```{r Load data, message=FALSE}
options(warn = -1) #avoiding unnecessary warrning messages
honeyraw9802 <- read_lines("honeyraw_1998to2002.csv")
honeyraw9802 <- data.frame(honeyraw9802[-1:-4])
str(honeyraw9802)

honeyraw0307 <- read_lines("honeyraw_2003to2007.csv")
honeyraw0307 <- data.frame(honeyraw0307[-1:-76])
str(honeyraw0307)

honeyraw0812 <- read_lines("honeyraw_2008to2012.csv")
honeyraw0812 <- data.frame(honeyraw0812[-1:-67])
str(honeyraw0812) #structure of our data frame
```


Combine the 3 datasets into a list, then loop through each dataset by index to perform cleaning. At the end of the loop, each cleaned dataframe is saved into a list of 3 clean data frames.

```{r Cleaning, message=FALSE, warning=FALSE}

honeyrawlist <- list(honeyraw9802,honeyraw0307,honeyraw0812)
honeyfinallist <- list()
for(honey_idx in 1:length(honeyrawlist)) {
  honey <- honeyrawlist[[honey_idx]]
  honey <- honey %>%
# Each dataframe is still csv formatted. Use separate() to split column 
# into multiple ones by comma. Set variable names and remove irrelevant columns.
    separate(colnames(honey[1]), paste0("X", 1:9), sep = ",") %>%
    setNames(c("X1", "X2", "state", "numcol", "yieldpercol","totalprod",
               "stocks","priceperlb","prodvalue")) %>%
    select(-X1, -X2)
# Glimpse shows that variables are considered characters. Convert each 
# variable to the correct class. Numbers are manipulated to remove the 
# scales (ex. number of colonies is in 1,000s). Remove double quotes
# from state abbreviations in state column.
  honey$state <- gsub("\"", "", honey$state)
  honey$numcol <- as.integer(honey$numcol)*1000
  honey$yieldpercol <- as.integer(honey$yieldpercol)
  honey$totalprod <- as.integer(honey$totalprod)*1000
  honey$stocks <- as.integer(honey$stocks)*1000
  honey$priceperlb <- currency(as.numeric(honey$priceperlb))/100
  honey$prodvalue <- currency(as.numeric(honey$prodvalue))*1000
  honey <- honey %>%
# Remove irrelevant rows by using filter() to select rows to keep. 
# These are rows that don't have NA values in prodvalue column and 
# rows that don't have forward slashes in state column.
    filter(!is.na(prodvalue)&!grepl("/", honey$state))
  honeyfinallist[[honey_idx]] <- honey
  
honeyfinaldf <- bind_rows(honeyfinallist)
glimpse(honeyfinaldf)  #almost clean data frame
}
```

### Attributes of the final data frame

- numcol: Number of honey producing colonies. Honey producing colonies are the maximum number of colonies from which honey was taken during the year. It is possible to take honey from colonies which did not survive the entire year

- yieldpercol: Honey yield per colony. Unit is pounds

- totalprod: Total production (numcol x yieldpercol). Unit is pounds

- stocks: Refers to stocks held by producers. Unit is pounds

- priceperlb: Refers to average price per pound based on expanded sales. Unit is dollars.

- prodvalue: Value of production (totalprod x priceperlb). Unit is dollars.



```{r honeyfinal_df, message=FALSE, warning=FALSE}

# some state names are in full length instead of abbreviations like the other ones.
#Use match() to find full state names and replace with abbreviations.
matches_and_nas <- match(honeyfinaldf$state, state.name)
matches <- !is.na(matches_and_nas)
honeyfinaldf$state[matches] <- state.abb[matches_and_nas[matches]]

# Lastly, we want to add a year column to represent which year each value
# belongs to. Data for different years are stored in contiguous blocks and 
# states are listed in alphabetical order. Each year begins with state AL 
# and ends with state WY. Thus, when states rotate back from WY to AL we 
# know the data is for a new year. Knowing this, we can use rep() to append 
# year information to each row. 
ystart <- c(grep("AL", honeyfinaldf$state))
yend <- c(grep("WY", honeyfinaldf$state))
years <- 1998:2012
ynums <- yend - ystart + 1
honeyfinaldf$year <- rep(years, ynums)

head(honeyfinaldf,20)  #Ready for analysis data frame(tidy data)
```

# Distribution Analysis

In this section of the data science processes, we will try to acess how well or how badly is our data distributed throughout time. We will try to analayze and find the best distribution that fits our data.
Before jumping into all the cool R codes that prepares graphs, it's a good start to start defining what distribution of data is. 

What is distribution ?

The distribution of a statistical data set (or a population) is a listing or function showing all the possible values (or intervals) of the data and how often they occur. When a distribution of categorical data is organized, you see the number or percentage of individuals in each group. When a distribution of numerical data is organized, they're often ordered from smallest to largest, broken into reasonably sized groups (if appropriate), and then put into graphs and charts to examine the shape, center, and amount of variability in the data.

For more Reading on Distribtion Analysis

<http://www.dummies.com/education/math/statistics/what-the-distribution-tells-you-about-a-statistical-data-set/>

It is alway a good idea to start with a scatter plot, to observe some behaviors of the data.

```{r distribution1 , message=FALSE}
honeyfinaldf %>%
  ggplot(aes(x = as.numeric(year), y = yieldpercol)) +
   geom_point()+
    labs(y="Yield per Colonies ", x = "Year (1998-2012)")
```
The plot dosen't get us what we wanted, so let't try a boxplot grouped by year.

```{r distribution , message=FALSE}
honeyfinaldf %>%
  ggplot(aes(x = year, y = yieldpercol, group = year)) +
    geom_boxplot(aes(color=year)) +
    labs(y="Yield per Colonies ", x = "Year (1998-2012)")
```

According to the above plot 

- Over time, namely after 2004, the mean yeild per colony for each year became unstable, which shows that the   industry was struggling after the epidemic and trying different methods to solve the issue could sometimes    succeed and give us the right skewed outliers and sometimes fials and move the mean and median to the left.

- The data is more or less centerd, except for the years after 2004, where it was bouncing up and down.

- In every year after 2008, there were one or more outliers that strongly contribute a skewed distribution

Let us now see how the price was changing through the given period 
```{r distributi , message=FALSE}
honeyfinaldf %>%
  ggplot(aes(x = as.numeric(year), y = priceperlb)) +
   geom_point()+
    labs(y="Price per pound", x = "Year (1998-2012)")
```

As we anticipated, the price was growing uniformly. This could be either from inflation or some other factor, we will investigate more in depth.

## Correlation and analysis.

In this section we are trying to find relationships between the attributes, wich can lead us to pridiction based off of some factors. 

Let's explore! Let's see how the total production and price of honey in the US has been changing over time.

```{r productionVStime, message=FALSE}

honeyfinaldf$prodvalue <- honeyfinaldf$totalprod *honeyfinaldf$priceperlb #fixing the round off error

mean_prod <- length(unique(honeyfinaldf$year))
for(i in unique(honeyfinaldf$year)) {
  mean_prod[i - unique(honeyfinaldf$year)[1] + 1] <- mean(honeyfinaldf[honeyfinaldf$year == i,]$totalprod)
}

qplot(x = unique(honeyfinaldf$year), y = mean_prod, xlab = "Year", ylab = "Mean Production (lbs)", main ="Mean Production vs Year") +
  geom_smooth(method = "lm")

mean_pric <- length(unique(honey$year))
for(i in unique(honeyfinaldf$year)) {
  mean_pric[i - unique(honeyfinaldf$year)[1] + 1] <- mean(honeyfinaldf[honeyfinaldf$year == i,]$priceperlb)
}
qplot(x = unique(honeyfinaldf$year), y = mean_pric, xlab = "Year", ylab = "Mean Price Per Pound", main ="Mean Price vs Year") +
  geom_smooth(method = "lm")
```

So far we know about the price and production changes through the given period of time, now lets see how heach state did on production and price in the year of 2012.

```{r honeyProduction, message=FALSE}
ggplot(data = honeyfinaldf[honeyfinaldf$year == 2012,], mapping = aes(x = state, y = priceperlb, fill = state)) +
  geom_bar(stat = "identity") +
  guides(fill = FALSE) + 
  xlab("State ") + 
  ylab("Honey price per pound") +
  ggtitle("Honey Price per pound in 2012") +
  coord_flip()

ggplot(data = honeyfinaldf[honeyfinaldf$year == 2012,], mapping = aes(x = state, y = totalprod, fill = state)) +
  geom_bar(stat = "identity") +
  guides(fill = FALSE) + 
  xlab("State ") + 
  ylab("Total Honey Produced") +
  ggtitle("Honey Production in 2012") +
  coord_flip()
```

## Regression analysis

Linear regression is a linear approach to modelling the relationship between a scalar response (or dependent variable) and one or more explanatory variables (or independent variables). The case of one explanatory variable is called simple linear regression. For more than one explanatory variable, the process is called multiple linear regression.

We will now try to see if there is a relationship between a scalar responses to fit a regression model.

```{r linear regression, message=FALSE}
honeyfinaldf %>%
  ggplot(aes(x=year,y=priceperlb, color=yieldpercol))+
  geom_point()+
  geom_smooth(method = lm)
```

Since the regression line is linear, which is an indicator of a relationship between price per pound of honey, the time(years), and the yield per colony; we will use yieldpercol and year to pridict(regress) the price.

```{r linear model, message=FALSE}
modelprice <- lm(formula = priceperlb ~ year * yieldpercol,data = honeyfinaldf)
broom::tidy(modelprice)
```

According to the above linear model, we fail to reject the null hypothesis(there is no relationship between the pridictors and the price per pount). The p-values of all three interactions are less than 0.05, as a result our model will look something like this:
$$y = -0.024295 + 0.12210*X_1 + 0.87507*X_2 - 0.00044*X_2$$
Where 

$$X_1 \rightarrow year\\
X_2 \rightarrow yieldpercol \\
X_3 \rightarrow year:yieldpercol\\
y \rightarrow price per pound
$$

## Conclusion and Aknowledgment 

We can pridict if the price of honey would increase or decrease the following year(2013) using our dataset.
Based on our linear model we fitted, we can conclude the following 

- The price of honey per pound would be -0.02 if we keep the three pridictors constant, which makes sense becuase if our production is zero, there isn nothing to sell.

- If we keep everything but the year pridictor constant our model looks like $$y = -0.024295 + 0.12210*X_1$$ 
This means price of honey per pound would increase on averge by 12% per year. Reasonably if we have 0 yield the only factor becomes inflation or some other factor.

- If we keep everything but the yield per colony constant our model looks like $$y = -0.024295 + 0.87507*X_2$$
This shows the dependence of price on the yield per colony predictor.

### Aknowledgement

This tutorial is based off of but not limited to the notes and codes from the course Introduction to Data Science in the University of Maryland College Park(Spring2018).




